---
title: "Terrain Analyses in R - Maggie"
author: "Maggie Church"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc:  yes
    toc_depth:  3
    toc_float: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F, message=F)

library(sf)
library(terra)
library(whitebox)
library(elevatr)
library(mapview)
library(tidyverse)
library(tmap)
```

# Assignment

## Q1 Generate your own watershed

Using the above code as a starter, generate your own watershed. Try to only pick a watershed that is not too large in size (\~ \< 1000km2). Remember you may need to mess around with the Z level for get_elev_raster().

Get DEM for the area around a PPR wetland
```{r, warning=F, message=F}
# reset the global environment
rm(list=ls())

# create a new folder for data files
if(!file.exists('dataPPR')){
  dir.create('dataPPR')
}
```

Hmmm I could try to do this for my entire study area, but my lil mac would die
```{R, eval=F}
# read in PPR boundary shapefile. I got it from Grady Mann (https://www.sciencebase.gov/catalog/item/54aeaef2e4b0cdd4a5caedf1)
roi <- st_read('dataPPR/input/ppr_boundary/ppr_boundary.shp') %>%
  st_set_crs(4326) %>%
  st_transform(2163)

```

### Calculate TWI in ND
Let's just compare a couple smaller areas across the region
```{r, warning=F, message=F}
#Create a pond point from Google Maps, using lat, long. 
NDpond <- tibble(site = c('ND'),
                long = c(48.734165),
                lat = c(-102.655226)) %>%
  st_as_sf(., coords = c('lat','long'), crs = 4263) %>%
  st_transform(2163)

#Save the point
st_write(NDpond, 'dataPPR/NDpond.shp', append = F)

#Use elevatr to download data and convert to terra object
elevND <- get_elev_raster(NDpond, z = 13) %>%
  rast(.)

#Visualize DEM
mapview(elevND) + mapview(NDpond)

#Save local version for whitebox to use
writeRaster(elevND, 'dataPPR/elevND.tif', overwrite = T)
```

Get flow accumulation using just the straight DEM. We won't
hydrologically condition bc we know our landscape isn't hydrologically connected. Also let's try DInf, which uses a multiple-flow direction. (**Note: I compared dInf to d8 and they produce the same TWI in the end...**)
```{r, warning=F, message=F}
#Get flow accumulation
wbt_d_inf_flow_accumulation(input = "dataPPR/elevND.tif",
                            output = "dataPPR/faND.tif")

#Read in fa
faND <- rast('dataPPR/faND.tif') %>%
  log10(.)

#visualize flow accumulation
mapview(faND) + mapview(NDpond)
```

Calculate TWI
```{r, warning=F, message=F}
#Get slope
wbt_slope(dem = 'dataPPR/elevND.tif',
          output = 'dataPPR/slopeND.tif',
          units = 'degrees')

#Generate TWI 
wbt_wetness_index(sca = 'dataPPR/faND.tif',
                  slope = 'dataPPR/slopeND.tif',
                  output = 'dataPPR/twiND.tif')

#Read in TWI and slope
twiND <- rast('dataPPR/twiND.tif')
slopeND <- rast('dataPPR/slopeND.tif')

#Visualize TWI
mapview(twiND) + mapview(NDpond)
```

### Calculate TWI in IA
Now run it back for an IA pond
```{r, warning=F, message=F}
#Create a pond point from Google Maps, using lat, long. 
IApond <- tibble(site = c('IA'),
                long = c(43.270079),
                lat = c(-94.868932)) %>%
  st_as_sf(., coords = c('lat','long'), crs = 4263) %>%
  st_transform(2163)

#Save the point
st_write(IApond, 'dataPPR/IApond.shp', append = F)

#Use elevatr to download data and convert to terra object
elevIA <- get_elev_raster(IApond, z = 13) %>%
  rast(.)

#Save local version for whitebox to use
writeRaster(elevIA, 'dataPPR/elevIA.tif', overwrite = T)

#Get flow accumulation
wbt_d_inf_flow_accumulation(input = "dataPPR/elevIA.tif",
                            output = "dataPPR/faIA.tif")

#Get slope
wbt_slope(dem = 'dataPPR/elevIA.tif',
          output = 'dataPPR/slopeIA.tif',
          units = 'degrees')

#Generate twi 
wbt_wetness_index(sca = 'dataPPR/faIA.tif',
                  slope = 'dataPPR/slopeIA.tif',
                  output = 'dataPPR/twiIA.tif')

#Read in TWI and slope
twiIA <- rast('dataPPR/twiIA.tif')
slopeIA <- rast('dataPPR/slopeIA.tif')

mapview(twiIA) + mapview(IApond)
```

### Compare
Print DEM resolutions, for context
```{r, warning=F, message=F}
NDres <- (cos(48.734165* pi/180) * 2 * pi * 6378137) / (256 * 2^13) 
IAres <- (cos(43.270079* pi/180) * 2 * pi * 6378137) / (256 * 2^13)
```
Resolution of ND DEM: `r round(NDres,2)`m 
Resolution of IA DEM: `r `round(IAres,2)`m

I know the bounding boxes are arbitrary here, but let's get a general idea of the average TWI in these two areas
```{r, warning=F, message=F}
# mean TWI for ND
global(twiND, fun = 'mean', na.rm = T)

# median TWI for ND
values(twiND) %>% median(na.rm=T)

# mean TWI for IA
global(twiIA, fun = 'mean', na.rm = T)

# median TWI for IA
values(twiIA) %>% median(na.rm=T)
```

Central measures for TWI are pretty darn similar between the two regions! Our IA region might be a bit wetter than our ND region.

### Plot TWI class separation using ground truth points
Ok for fun, I'll import my ground-truth points and see how their TWIs compare. 
```{R, warning=F, message=F}
file_list <- list.files("dataPPR/input/GTPoints", pattern = "*shp", full.names = TRUE)[-2]
shapefile_list <- lapply(file_list, read_sf)

wetGT <- do.call(bind_rows, shapefile_list) %>%
  st_set_crs(4326) %>%
  st_transform(2163)

nonwetGT <- st_read('dataPPR/input/GTPoints/nonwetGTPoints.shp') %>%
  st_set_crs(4326) %>%
  st_transform(2163)
```

Extract TWI to GT points and plot
```{r, warning=F, message=F}

# keep GT points that fall within that ROI
wetGTND <- crop(vect(wetGT), ext(twiND))
nonwetGTND <- crop(vect(nonwetGT), ext(twiND))

# extract TWI values to GT points
wetGTND_wTWI <- terra::extract(twiND, wetGTND, bind=F) %>% mutate(class='wet')
nonwetGTND_wTWI <- terra::extract(twiND, nonwetGTND, bind=F) %>% mutate(class='nonwet')

### Do the same for ND ###

# keep GT points that fall within that ROI
wetGTIA <- crop(vect(wetGT), ext(twiIA))
nonwetGTIA <- crop(vect(nonwetGT), ext(twiIA))

# extract TWI values to GT points
wetGTIA_wTWI <- terra::extract(twiIA, wetGTIA, bind=F) %>% mutate(class='wet')
nonwetGTIA_wTWI <- terra::extract(twiIA, nonwetGTIA, bind=F) %>% mutate(class='nonwet')

# Plot distribution of TWI in ND
bind_rows(wetGTND_wTWI, nonwetGTND_wTWI) %>% 
  filter(!is.na(twiND)) %>%
  ggplot(aes(x=twiND, y=class)) +
  geom_boxplot()+
  labs(title="North Dakota roi")

# Plot distribution of TWI in IA
 bind_rows(wetGTIA_wTWI, nonwetGTIA_wTWI) %>% 
  filter(!is.na(twiIA)) %>%
  ggplot(aes(x=twiIA, y=class)) +
  geom_boxplot() +
  labs(title="Iowa roi")
 
 avSlopeIA <- global(slopeIA, fun = 'mean', na.rm = T)
 avSlopeND <- global(slopeND, fun = 'mean', na.rm = T)
``` 

What the heck? Well the distinction between land cover classes is strong in ND, that's good. But why do these Iowa wetlands have a lower TWI than the nonwetlands? Maybe the distinction in IA isn't as strong because this area is flatter? The average slope in the ND area is `r round(avSlopeND,2)` and the average slope in the IA area is `r round(avSlopeIA, 2)`.

### Plot HAND class separation using ground truth points
One last thing. I just learned about HAND (height above nearest drainage), aka elevation above stream. Let's plot that bad boy too. I read that you can use permanent riverine features from the National Hydrography Dataset as the stream input for HAND, but I'm gonna take a lazier approach for now.
```{R, warning=F, message=F}
# extract likely stream cells from fa
wbt_extract_streams(
  flow_accum = 'dataPPR/faND.tif',
  output = 'dataPPR/streamsND.tif',
  threshold = 10^2.5 #arbitrary, based on map
)

# calculate HAND
wbt_elevation_above_stream(
    dem = 'dataPPR/elevND.tif',
    streams = 'dataPPR/streamsND.tif',
    output = 'dataPPR/handND.tif'
)

# visualize
handND <- rast('dataPPR/handND.tif')
mapview(handND) + mapview(NDpond)

#### Iowa's turn ###
# extract likely stream cells from fa
wbt_extract_streams(
  flow_accum = 'dataPPR/faIA.tif',
  output = 'dataPPR/streamsIA.tif',
  threshold = 10^2.5 #arbitrary, based on map
)

# calculate HAND
wbt_elevation_above_stream(
    dem = 'dataPPR/elevIA.tif',
    streams = 'dataPPR/streamsIA.tif',
    output = 'dataPPR/handIA.tif'
)

# visualize
handIA <- rast('dataPPR/handIA.tif')
mapview(handIA) + mapview(IApond)
```
A lot of ponds got converted to NA merp.

Extract HAND to GT points and plot
```{r, warning=F, message=F}
# extract HAND values to GT points
wetGTIA_wHAND <- terra::extract(handIA, wetGTIA, bind=F) %>% mutate(class='wet')
nonwetGTIA_wHAND <- terra::extract(handIA, nonwetGTIA, bind=F) %>% mutate(class='nonwet')

# extract HAND values to GT points
wetGTND_wHAND <- terra::extract(handND, wetGTND, bind=F) %>% mutate(class='wet')
nonwetGTND_wHAND <- terra::extract(handND, nonwetGTND, bind=F) %>% mutate(class='nonwet')

# Plot distribution of HAND in ND
bind_rows(wetGTND_wHAND, nonwetGTND_wHAND) %>% 
  filter(!is.na(handND)) %>%
  ggplot(aes(x=handND, y=class)) +
  geom_boxplot()+
  labs(title="North Dakota roi")

# Plot distribution of HAND in IA
 bind_rows(wetGTIA_wHAND, nonwetGTIA_wHAND) %>% 
  filter(!is.na(handIA)) %>%
  ggplot(aes(x=handIA, y=class)) +
  geom_boxplot() +
  labs(title="Iowa roi")
```
Whoa that's super unexpected. ND looks super solid. Why is IA so weird??

